---
title: "Ordination of maunga on GBI for JO"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
library(tidyverse)
library(patchwork)
library(ggrepel)
# library(ggvegan)
library(ggdendro)
library(dendextend)
library(sf)
library(ggspatial)

source('helperPlots.r')
# here::here()
```


```{r load_data, echo = FALSE}
mts.woody <- read.csv('../data/woodyAltitude_rev.csv', header = T, row.names = 1)
mts.ferns <- read.csv('../data/fernAltitude_rev.csv', header = T, row.names = 1)

# get the site codes and elevations
site <- str_sub(names(mts.woody), 1, 1)
elev <- as.numeric(str_sub(names(mts.woody), 3, 5))

# transpose and convert NAs to zero
mts.woody <- t(mts.woody)
mts.woody[is.na(mts.woody)] <- 0

mts.ferns <- t(mts.ferns)
mts.ferns[is.na(mts.ferns)] <- 0
```

## Map of GBI

Figure 2 = map (to be tidied)

```{r map, echo = FALSE, message = FALSE}
coast <- read_sf("../data/nz_coast_NZTM")

bd <- c(1741620, 5859550, 1834360,	6012920)
coast.gbi <- st_crop(coast, xmin = 1741620, ymin = 5859550, xmax = 1834360,	ymax = 6012920)

places <- read.csv('../data/map_places.csv')

ggplot(data = coast.gbi) +
  geom_sf(col = 'blue') +
  geom_point(data = places, aes(x = long.tm, y = lat.tm)) +
  geom_text_repel(data = places, aes(x = long.tm, y = lat.tm, label = name), size = 4.5, alpha = 0.7) +
  xlab('Latitude') +
  ylab('Longitude') + 
  annotation_scale(location = "bl", width_hint = 0.4) +
  theme_bw()
```


## Summary of species richness trends with elevation by site for the woody and ferns

This will be Figure 3

```{r richness, echo = FALSE, message = FALSE}
all.data <- data.frame(elev = elev, site = site, woody = rowSums(mts.woody), ferns = rowSums(mts.ferns))

all.data.long <- all.data %>%
  pivot_longer(cols = 3:4, values_to = 'richness', names_to = 'taxa')

elev.s.woody <- ggplot(filter(all.data.long, taxa == 'woody')) +
  geom_point(aes(x = elev, y = richness, col= site, shape = site), size = 2.5) +
  scale_colour_brewer(type = 'qual', name = "Maunga", labels = c("Hirikimata", "Ruahine", "Tataweka")) +
  scale_shape_manual(values = c(16,17, 15), name = "Maunga", labels = c("Hirikimata", "Ruahine", "Tataweka")) +  
  labs(x = 'Elevation (m)', y = 'Species richness') +
  geom_smooth(aes(x = elev, y = richness), method = 'lm') +
  ylim(0,70) +
  theme_minimal()

elev.s.ferns <- ggplot(filter(all.data.long, taxa == 'ferns')) +
  geom_point(aes(x = elev, y = richness, col= site, shape = site), size = 2.5) +
  scale_colour_brewer(type = 'qual', name = "Maunga", labels = c("Hirikimata", "Ruahine", "Tataweka")) +
  scale_shape_manual(values = c(16,17, 15), name = "Maunga", labels = c("Hirikimata", "Ruahine", "Tataweka")) +  
  labs(x = 'Elevation (m)', y = 'Species richness') +
  geom_smooth(aes(x = elev, y = richness), method = 'lm') +
  ylim(0,70) +
  theme_minimal() +
  theme(axis.title.y = element_blank())
  
ee <- elev.s.woody + elev.s.ferns +
  plot_annotation(tag_levels = 'A') + plot_layout(guides = 'collect')
ee & theme(legend.position = 'bottom')
  
```

## Pairwise differences  of sites (different mt) 

This will be Figure 4

```{r dist_mtx, echo = FALSE}
# bray-curtis distance matrix
woody.dist <- vegdist(mts.woody, method = 'bray')
ferns.dist <- vegdist(mts.ferns, method = 'bray')
```


```{r woody_pairwise, echo = FALSE}
woody.mts.dist.df <- data.frame(as.matrix(woody.dist)) %>%
  rownames_to_column(var = 'site_elev') %>%
  pivot_longer(cols = -1) %>%
  separate(site_elev, c("site1", "elev1")) %>%
  separate(name, c("site2", "elev2")) %>%
  distinct() %>%
  mutate(same_elev = ifelse(elev1 == elev2 & site1 != site2 , 1, 0)) %>%
  mutate(site_pair = paste0(site1, '.', site2))
  
woody.pairwise.bc <- woody.mts.dist.df %>%
  filter(same_elev == 1) %>%
  filter(site_pair == 'R.H' | site_pair == 'T.H' | site_pair == 'T.R')

woody.pairs.gg <- ggplot(woody.pairwise.bc) +
  geom_point(aes(y = value, x = elev1, col = site_pair, shape = site_pair), size = 3) +
  labs(x = 'Elevation (m)', y = 'Dissimilarity (Bray-Curtis)') +
  scale_colour_brewer(type = 'qual', name = 'Site pair') +
  ylim(0,1) +
  guides(shape = FALSE) +
  theme_bw()
```

also

```{r ferns_pairwise, echo = FALSE}
# Make a long version of the distance matrix
ferns.mts.dist.df <- data.frame(as.matrix(ferns.dist)) %>%
  rownames_to_column(var = 'site_elev') %>%
  pivot_longer(cols = -1) %>%
  separate(site_elev, c("site1", "elev1")) %>%
  separate(name, c("site2", "elev2")) %>%
  distinct() %>%
  mutate(same_elev = ifelse(elev1 == elev2 & site1 != site2 , 1, 0)) %>%
  mutate(site_pair = paste0(site1, '.', site2))

ferns.pairwise.bc <- ferns.mts.dist.df %>%
  filter(same_elev == 1) %>%
  filter(site_pair == 'R.H' | site_pair == 'T.H' | site_pair == 'T.R')

ferns.pairs.gg <- ggplot(ferns.pairwise.bc) +
  geom_point(aes(y = value, x = elev1, col = site_pair, shape = site_pair), size = 3) +
  labs(x = 'Elevation (m)', y = 'Dissimilarity (Bray-Curtis)') +
  scale_colour_brewer(type = 'qual', name = 'Site pair') +
  ylim(0,1) +
  guides(shape = FALSE) +
  theme_bw() +
  theme(axis.title.y = element_blank())
```


```{r, echo = FALSE}
pp <- woody.pairs.gg + ferns.pairs.gg + plot_annotation(tag_levels = 'A') + plot_layout(guides = 'collect')
pp & theme(legend.position = 'bottom') 
```


## Hierarchical classifications for the woody and ferns

This will be Figure 5

```{r hac, echo = FALSE}

pal <- RColorBrewer::brewer.pal(3, "Accent")   

woody.hcl <- hclust(woody.dist, method = 'average')
woody.hcl <- as.dendrogram(woody.hcl)

# sort sites based on appearance in the dendrogram

woody.hcl.col <- as.numeric(factor(site[order.dendrogram(woody.hcl)]))

woody.hcl <- woody.hcl %>% 
set("labels_colors", pal[woody.hcl.col]) %>%
set("leaves_pch", 19) %>%
set("leaves_col", pal[woody.hcl.col]) %>%
set("leaves_cex", 2)

woody.hcl.gg <- as.ggdend(woody.hcl) %>% 
ggplot() + 
scale_x_discrete(labels = NULL, breaks = NULL) + 
labs(x = "", y = 'Distance') + 
theme_minimal()

## Now the ferns
ferns.hcl <- hclust(ferns.dist, method = 'average')

ferns.hcl <- as.dendrogram(ferns.hcl)
ferns.hcl.col <- as.numeric(factor(site[order.dendrogram(ferns.hcl)]))


ferns.hcl <- ferns.hcl %>% 
set("labels_colors", pal[ferns.hcl.col]) %>%
set("leaves_pch", 19) %>%
set("leaves_col", pal[ferns.hcl.col]) %>%
set("leaves_cex", 2)

ferns.hcl.gg <- as.ggdend(ferns.hcl) %>% 
ggplot() + 
scale_x_discrete(labels = NULL, breaks = NULL) +
 labs(x = "", y = 'Distance') + 
 theme_minimal()

woody.hcl.gg + ferns.hcl.gg + plot_annotation(tag_levels = 'a')
```

## Ordination (nMDS) for the woody species

This will be figure 6

```{r nmds, echo = FALSE, message = FALSE}
# nMDS
woody.mds <- metaMDS(mts.woody, distance = 'bray', wascores = TRUE, autotransform = FALSE, k = 2, trace = 0)

#plot(woody.mds, type = 't', display = 'sites')
woody.mds.sc <- data.frame(scores(woody.mds), site.code = rownames(mts.woody), site = site, elev = elev)

woody.stress.lbl <- paste('Stress = ', round(woody.mds$stress, 4))
# plot.mds.gg <- function(mds, txt.col = 'blue', txt.x = 0, txt.y = 0, clusters = NULL, labels = FALSE)

woody.mds.simp.gg <- plot.mds.gg(woody.mds, clusters = site, txt.x = 1, txt.y = 0.7) +
  geom_text_repel(data = woody.mds.sc, aes(x = NMDS1, y = NMDS2, label = site.code)) +
  scale_colour_brewer(type = 'qual', name = "Maunga", labels = c("Hirikimata", "Ruahine", "Tataweka")) +
  guides(shape = FALSE) +
  theme(legend.position = "bottom")
```

```{r, echo = FALSE}
woody.mds.simp.gg
```


## Ordination (nMDS) for the ferns species

This will be figure 7

```{r ferns_nmds, echo = FALSE, message = FALSE}
# nMDS
ferns.mds <- metaMDS(mts.ferns, distance = 'bray', wascores = TRUE, autotransform = FALSE, k = 2, rty = 100, trace = 0)

#plot(ferns.mds, type = 't', display = 'sites')
ferns.mds.sc <- data.frame(scores(ferns.mds), site.code = rownames(mts.ferns), site = site, elev = elev)

ferns.stress.lbl <- paste('Stress = ', round(ferns.mds$stress, 4))

ferns.mds.simp.gg <- plot.mds.gg(ferns.mds, clusters = site, txt.x = 0.7, txt.y = -1) +
  geom_text_repel(data = ferns.mds.sc, aes(x = NMDS1, y = NMDS2, label = site.code)) +
  scale_colour_brewer(type = 'qual', name = "Site", labels = c("Hirikimata", "Ruahine", "Tataweka")) +
  guides(shape = FALSE) +
  theme(legend.position = "bottom")
```

```{r, echo = FALSE}
ferns.mds.simp.gg
```


## Range barplots

```{r extract_names, echo = FALSE}
# Woody spp
mts.woody.df <- data.frame(mts.woody)
mts.woody.df <- mts.woody.df %>%
mutate(site_synth = ifelse(site == 'H', 'H', 'T_R')) %>%
rownames_to_column(var = 'site_elev') %>%
separate(site_elev, into = c('site', 'elev'))

mts.woody.df <- mts.woody.df %>% select(site, aleexc:weisil)

mts.split <- mts.woody.df %>% split(mts.woody.df$site)

mts.split$H <- mts.split$H[,-(1:2)]
mts.split$T <- mts.split$T[,-(1:2)]
mts.split$R <- mts.split$R[,-(1:2)]

nh <- colnames(mts.split$H[, colSums(mts.split$H) > 0])
nt <- colnames(mts.split$T[, colSums(mts.split$T) > 0])
nr <- colnames(mts.split$R[, colSums(mts.split$R) > 0])

spp.all <- unique(c(nh, nt, nr))  # species list for all three sites
woody.all_sites <- Reduce(intersect, list(nh, nt, nr)) # spp present at all three sites
woody.hira <- nh
woody.hira_only <- Reduce(setdiff, list(nh, nt, nr)) # spp only at Hira

# Species present at all three sites at 150 m
lowl.spp <- colSums(mts.woody.df[elev == 150,-1]) == 3 & colSums(mts.woody.df[elev == 400]) == 0

woody.all_sites_150 <- names(mts.woody.df)[-1][]

########################
# Fern spp
mts.ferns.df <- data.frame(mts.ferns)
mts.ferns.df$site <- site

mts.ferns.df <- mts.ferns.df %>% select(site, adicun:tristr)

mts.fern.split <- split(mts.ferns.df, mts.ferns.df$site)

mts.fern.split$H <- mts.fern.split$H[,-(1:2)]
mts.fern.split$T <- mts.fern.split$T[,-(1:2)]
mts.fern.split$R <- mts.fern.split$R[,-(1:2)]

nh.f <- colnames(mts.fern.split$H[, colSums(mts.fern.split$H) > 0])
nt.f <- colnames(mts.fern.split$T[, colSums(mts.fern.split$T) > 0])
nr.f <- colnames(mts.fern.split$R[, colSums(mts.fern.split$R) > 0])

spp.ferns.all <- unique(c(nh.f, nr.f, nt.f))
ferns.all_sites <- Reduce(intersect, list(nh.f, nr.f, nt.f)) # spp in all three
ferns.hira <- nh.f
ferns.hira_only <- Reduce(setdiff, list(nh.f, nt.f, nr.f)) # spp only at Hira
```

```{r extract_range_data, echo = FALSE}

# Hirakimata ONLY
hira_only.range <- extract.limits(mts.woody, woody.hira_only, elev)
hira_only.ferns.range <- extract.limits(mts.ferns, ferns.hira_only, elev)

# Species at Hirakimata and elsewhere
# Using grep to ignore R and T
hira.set <- grep(pattern = 'H', rownames(mts.woody))

hira.range <- extract.limits(mts.woody[hira.set,], woody.hira, elev) 
hira.ferns.range <- extract.limits(mts.ferns[hira.set,], ferns.hira, elev)

## Present at ALL THREE mts (sites)
all_sites.range <- extract.limits(mts.woody, woody.all_sites, elev)
all_sites.ferns.range <- extract.limits(mts.ferns, ferns.all_sites, elev)
```

```{r build_range_plots, echo = FALSE}
pal <- RColorBrewer::brewer.pal(3, "Accent")   

# ordering by min
hira.wdy.rg <- ggplot(data = hira_only.range$elev.lims) +
  geom_segment(aes(x = min, xend = max, y = spp, yend = spp), size = 3) +
  geom_point(data = hira_only.range$elev.lims.sgl, aes(x = min, y = spp), shape = 3) +
  labs(x = "Elevation (m)", y = "Species") +
  theme_bw()

hira.fern.rg <- ggplot(data = hira_only.ferns.range$elev.lims) +
  geom_segment(aes(x = min, xend = max, y = spp, yend = spp), size = 3) +
  geom_point(data = hira_only.ferns.range$elev.lims.sgl, aes(x = min, y = spp), shape = 3) +
  labs(x = "Elevation (m)", y = "Species") +
  theme_bw()

# Extract the spp at Hira also present elsewhere (as may have different range) - in both cases have to be at 150 m
hira.e <- hira.range$elev.lims[str_to_lower(hira.range$elev.lims$spp) %in% woody.all_sites_150, ]
all.e <- all_sites.range$elev.lims[str_to_lower(all_sites.range$elev.lims$spp) %in% woody.all_sites_150, ]

# Plots ordered by upper limit on Hirikimata
hira.e$spp <- fct_reorder(hira.e$spp, hira.e$max, min)
all.e$spp <- fct_reorder(all.e$spp, hira.e$max, min)

# Spp at just one site at H
hira.sgl.e <- filter(hira.e, single == 1)

all.wdy.rg <- ggplot() +
  geom_segment(data = all.e, aes(x = min, xend = max, y = spp, yend = spp), size = 1.5, col = 'grey') +
  geom_segment(data = hira.e, aes(x = min, xend = max, y = spp, yend = spp), size = 1.5,  col = pal[1]) +
  geom_point(data = hira.sgl.e, aes(y = spp, x = min), size = 2,  col = pal[1]) +
  labs(x = "Elevation (m)", y = "Species") +
  theme_bw()

# And now the fern taxa
hira.f.e <- hira.ferns.range$elev.lims[str_to_lower(hira.ferns.range$elev.lims$spp) %in% ferns.all_sites, ]
all.f.e <- all_sites.ferns.range$elev.lims

hira.f.e$spp <- fct_reorder(hira.f.e$spp, hira.f.e$max, min)
all.f.e$spp <- fct_reorder(all.f.e$spp, hira.f.e$max, min)  

# Spp at just one site at H
hira.sgl.f.e <- filter(hira.f.e, single == 1)

all.fern.rg <- ggplot() +
  geom_segment(data = all.f.e, aes(x = min, xend = max, y = spp, yend = spp), size = 1.5, col = 'grey') +
  geom_segment(data = hira.f.e, aes(x = min, xend = max, y = spp, yend = spp), size = 1.5,  col = pal[1]) +
  labs(x = "Elevation (m)", y = "Species") +
  geom_point(data = hira.sgl.f.e, aes(y = spp, x = min), size = 2, col = pal[1]) +
  theme_bw()
```

Plots of the ranges of (a) woody species and (b) ferns on Hirakimata _only_ and all three sites at an elevation of 150 m to 400 m (lower bnd, upper bnd) -> 'lowland' species

```{r plot_ranges, echo = FALSE}
library(patchwork)

hira.wdy.rg | hira.fern.rg +
  plot_annotation(tag_levels = 'a')


all.wdy.rg | all.fern.rg +
  plot_annotation(tag_levels = 'a')
```
